Phenotypic evolution is more constrained in simpler food webs

### 2019-03-29 manuscript.Rmd
edited discussion to reduce emphasis on food-web complexity.

<!--Summary-->
We found that simpler food webs constrained phenotypic evolution in gall midges in two key ways. First, more traits contributed to the slope of the adaptive landscape in the simpler food web, suggesting greater constraints on the trajectory of phenotypic evolution. Second, the simple food-web treatment indirectly increased the number of genetic constraints, which could act to constrain the gall midge's adaptive potential in the face of novel selection pressures. We also found evidence of indirect selection pressures in more complex food webs, suggesting that the adaptive landscape may be more dynamic in complex food webs. Taken together, our study provides experimental evidence from the field that the simplification of ecological communities may constrain the adaptive potential of remaining populations.

<!--Selective constraints-->
All three phenotypic traits we examined experienced directional selection in the simpler food web, indicating that there is an optimal phenotype that maximizes larval survival (i.e. large chamber diameter, small clutch size, and avoidance of conspecifics). In contrast, we did not observe selective constraints on clutch size and oviposition preference in the more complex food web. This suggests that there are many optimal phenotypes (adaptive peaks), or in this case, a flat adaptive landscape where there are no fitness consequences for phenotypic change in these traits. This also implies that as selective constraints dampen in more complex food webs, then the trajectory of evolution becomes more determined by genetic constraints. This is made clear if we inspect the 'Lande equation', $\Delta z = \text{G} \beta$, where G corresponds to the population's G-matrix. As the components of $\beta$ go to zero, then $\Delta z$ is more influenced by the G-matrix. Interestingly, this aligns with @Schluter1996's hypothesis that phenotypic evolution often follows 'genetic lines of least resistance'. @Schluter1996 found support for this hypothesis from data on natural populations of several vertebrate species, including threespine sticklebacks, a few species of songbirds, and mice from the genus *Peromyscus*. All of these species occupy intermediate trophic levels and are likely embedded in complex food webs, which is consistent with our suggestion that genetic constraints may have a stronger influence in more complex food webs.

<!--Indirect effects of selective constraints-->
We also found evidence for more genetic constraints in simpler food webs due to indirect effects of selection on the population's G-matrix. The ability of a population to adapt to novel selection pressures (evolvability) is largely governed by the structure of its G-matrix [@Hansen2008]. When selection favors genetic covariance between traits (positive or negative), this results in less autonomy of evolutionary responses to changing environments. Similarly, decreases in genetic variance constrain potential for the trait itself to evolve. Together, this suggests that simpler food webs may decrease the evolutionary potential of populations by indirectly selecting for decreases in genetic variance in multiple traits and favoring trait integration. Current theory often assumes genetic variances and covariances remain constant over time rather than dynamically changing with the community context [@McPeek2017; @Guimaraes2017]. Our empirical results highlight the need to explore the evolutionary consequences of not only direct effects of selection, but indirect effects on genetic constraints that are shaped by the community context.

<!-- Generality of results -->
The generality of our results likely depends on the relative abundance and functional differences between consumers in a community. For example, if consumers do not differ from each other, then we do not expect changes in food-web complexity to modify selective constraints. Also, many consumers may be at too low of abundances to impose selection on their resources. Rank abundance curves [@Preston1948] and the disproportionate number of weak interactions in diverse communities [@Paine1992] support this notion. When consumers are abundant though, the effect of food-web complexity will depend on whether different species impose conflicting selection pressures or select for distinct traits. For example, parisitoids and birds impose conflicting selection pressures on the size of galls induced by the fly *Eurosta solidaginis* [@Weis1985; @Abrahamson1997]. Recent studies in this system have shown that decreases in the relative abundance of birds, due to either small patch sizes [@Start2016] or proximity to urban areas [@Start2018] causes a shift from neutral to directional selection on gall size. On the other hand, different consumers may impose selection on different traits, favoring trait integration and increasing genetic constraints. Examples of this include strong genetic covariances in plant resistance to different insect herbivores [@Maddox1990; @Wise2007; @Wise2013], although there are also examples where these covariances are weak [@Roche1997; @Barbour2015], or vary from year-to-year [@Johnson2007]. We suggest that gaining predictive insight to the evolutionary consequences of food-web disassembly requires an understanding of the mechanisms governing the assembly of trophic interactions. <!--We suggest that gaining predictive insight to the effects of food-web disassembly requires an understanding of the mechanisms governing the assembly of trophic interactions.-->

<!--Indirect evolutionary effects-->
Our results suggest that the simplification of ecological communities may not only directly affect connected species, but also result in indirect evolutionary effects. In our study, this indirect effect arises from egg parasitoids being released from intraguild predation in simpler food webs. This release occurs more on trees with high larval densities, which could intensify selection on gall midge oviposition preference. A growing number of experiments over the past two decades have demonstrated the presence and potential importance of indirect evolutionary effects that emerge in a community context [@Juenger1998; @Stinchcombe2001; @Lankau2007; @Walsh2008; @Walsh2010; @terHorst2010; @Sahli2011; @Lau2012; @terHorst2015; @Schiestl2018; @StartInPress]. If indirect evolutionary effects are common [@Walsh2013], then predicting evolutionary trajectories resulting from the simplification of food webs will require evolutionary studies to explicitly account for the ecological networks that species are embedded in.

<!-- Community context shaping adaptation to environmental change --> 
There is a growing number of theoretical studies on adaptation to directional environmental change that incorporate species interactions [e.g. @deMazancourt2008; @Johansson2008; @Norberg2012; @Osmond2017PredatorsHelpPrey]. This work has given insight to the mechanisms by which pairwise interactions modify the adaptive potential of species. Our study hints at a novel mechanism, whereby food-web complexity flattens the adaptive landscape, thus facilitating future adaptation by allowing genetic and phenotypic variation to persist. This mechanism only emerges once we move beyond pairwise interactions to consider selection on multiple traits in a community context. However, a study on competition has highlighted that we may expect the opposite effect of species diversity in competitive communities [@deMazancourt2008]. This negative effect of species diversity on adaptation occurs because there is a greater chance that the optimal phenotype is already occupied by a competitor in a more diverse community. More work is needed to examine how the distribution of different interaction types affects adaptation to environmental change in species-rich communities.

<!-- Conclusions -->
Our study gives insight to how local extinctions, and concomitant decreases in food-web complexity, increase evolutionary constraints. This hints at a potential insidious effect of local extinctions that compromises the robustness of remaining populations to future environmental change. Our work also highlights some key challenges for predicting phenotypic evolution in the face of global change. First, the simplification of ecological communities may actually reduce the predictability of phenotypic evolution, because knowledge is required of both selective and genetic constraints, rather than potentially just genetic constraints in more complex systems. Second, many theoretical models of eco-evolutionary dynamics focus on phenotypic change in a single trait, yet our results highlight that the number of traits under selection may change with the community context. Importantly, while the net effects of selection in complex communities may be dampened, we still found that different species/guilds imposed different selection pressures. Knowing these hidden selection pressures is critical for prediction, because the trajectory of evolution will depend on the nature of change in the community context. We expect that a continued integration of adaptive landscapes and ecological networks will enhance our ability to predict the evolutionary consequences of changes in ecological communities.

>>>>

<!--Summary-->
We found that the extinction of larval parasitoids constrained phenotypic evolution in gall midges in two key ways. First, more traits contributed to the slope of the adaptive landscape in the absence of larval parasitoids, suggesting greater constraints on the trajectory of phenotypic evolution. Second, excluding larval parasitoids indirectly increased the number of genetic constraints, which could act to constrain the gall midge's adaptive potential in the face of novel selection pressures. Our experiment also revealed evidence of indirect selection pressures, suggesting that the loss of species interactions may have complex effects on the trajectories of phenotypic evolution. Taken together, our study provides experimental evidence from the field that the loss of species interactions may constrain the adaptive potential of remaining populations.

<!--Selective constraints-->
All three phenotypic traits we examined experienced directional selection in the absence of larval parasitoids, indicating that there is an optimal phenotype that maximizes larval survival (i.e. large chamber diameter, small clutch size, and avoidance of conspecifics). In contrast, we did not observe selective constraints on clutch size and oviposition preference in the presence of larval parasitoids. This suggests that there are many optimal phenotypes (adaptive peaks), or in this case, a flat adaptive landscape where there are no fitness consequences for phenotypic change in these traits. 
<!-- need to see if I can salvage this, because as the editor points out, it is not exactly true.-->
This also implies that as selective constraints dampen in more complex food webs, then the trajectory of evolution becomes more determined by genetic constraints. This is made clear if we inspect the 'Lande equation', $\Delta z = \text{G} \beta$, where G corresponds to the population's G-matrix. As the components of $\beta$ go to zero, then $\Delta z$ is more influenced by the G-matrix. 
Interestingly, this aligns with @Schluter1996's hypothesis that phenotypic evolution often follows 'genetic lines of least resistance'. @Schluter1996 found support for this hypothesis from data on natural populations of several vertebrate species, including threespine sticklebacks, a few species of songbirds, and mice from the genus *Peromyscus*. All of these species occupy intermediate trophic levels and are likely embedded in complex food webs, which is consistent with our suggestion that genetic constraints may have a stronger influence in more complex food webs.

<!--Indirect effects of selective constraints-->
We also found evidence for more genetic constraints when we excluded larval parasitoids due to indirect effects of selection on the population's G-matrix. The ability of a population to adapt to novel selection pressures (evolvability) is largely governed by the structure of its G-matrix [@Hansen2008]. When selection favors genetic covariance between traits (positive or negative), this results in less autonomy of evolutionary responses to changing environments. Similarly, decreases in genetic variance constrain potential for the trait itself to evolve. Together, this suggests that the loss of species interactions may decrease the evolutionary potential of populations by indirectly selecting for decreases in genetic variance in multiple traits and favoring trait integration. Current theory often assumes genetic variances and covariances remain constant over time rather than dynamically changing with the community context [@McPeek2017; @Guimaraes2017]. Our empirical results highlight the need to explore the evolutionary consequences of not only direct effects of selection, but indirect effects on genetic constraints that are shaped by the community context.

<!-- Generality of results -->
The generality of our results likely depends on the relative abundance and functional differences between consumers in a community. For example, if consumers do not differ from each other, then we do not expect changes in food-web structure to modify selective constraints. Also, many consumers may be at too low of abundances to impose selection on their resources. Rank abundance curves [@Preston1948] and the disproportionate number of weak interactions in diverse communities [@Paine1992] support this notion. When consumers are abundant though, the effect of food-web structure will depend on whether different species impose conflicting selection pressures or select for distinct traits. For example, parisitoids and birds impose conflicting selection pressures on the size of galls induced by the fly *Eurosta solidaginis* [@Weis1985; @Abrahamson1997]. Recent studies in this system have shown that decreases in the relative abundance of birds, due to either small patch sizes [@Start2016] or proximity to urban areas [@Start2018] causes a shift from neutral to directional selection on gall size. On the other hand, different consumers may impose selection on different traits, favoring trait integration and increasing genetic constraints. Examples of this include strong genetic covariances in plant resistance to different insect herbivores [@Maddox1990; @Wise2007; @Wise2013], although there are also examples where these covariances are weak [@Roche1997; @Barbour2015], or vary from year-to-year [@Johnson2007]. We suggest that gaining predictive insight to the evolutionary consequences of food-web disassembly requires an understanding of the mechanisms governing the assembly of trophic interactions. <!--We suggest that gaining predictive insight to the effects of food-web disassembly requires an understanding of the mechanisms governing the assembly of trophic interactions.-->

<!--Indirect evolutionary effects-->
Our results suggest that the loss of species interactions may not only directly affect connected species, but also result in indirect evolutionary effects. In our study, this indirect effect arises from egg parasitoids being released from intraguild predation when we exlcuded larval parasitoids. This release occurs more on trees with high larval densities, which could intensify selection on gall midge oviposition preference. A growing number of experiments over the past two decades have demonstrated the presence and potential importance of indirect evolutionary effects that emerge in a community context [@Juenger1998; @Stinchcombe2001; @Lankau2007; @Walsh2008; @Walsh2010; @terHorst2010; @Sahli2011; @Lau2012; @terHorst2015; @Schiestl2018; @StartInPress]. If indirect evolutionary effects are common [@Walsh2013], then predicting evolutionary trajectories resulting from the loss of species interactions will require evolutionary studies to explicitly account for the ecological networks that species are embedded in.

<!-- Community context shaping adaptation to environmental change --> 
There is a growing number of theoretical studies on adaptation to directional environmental change that incorporate species interactions [e.g. @deMazancourt2008; @Johansson2008; @Norberg2012; @Osmond2017PredatorsHelpPrey]. This work has given insight to the mechanisms by which pairwise interactions modify the adaptive potential of species. Our study hints at a novel mechanism, whereby more complex food webs flatten the adaptive landscape, thus facilitating future adaptation by allowing genetic and phenotypic variation to persist. This mechanism only emerges once we move beyond pairwise interactions to consider selection on multiple traits in a community context. However, a study on competition has highlighted that we may expect the opposite effect of species diversity in competitive communities [@deMazancourt2008]. This negative effect of species diversity on adaptation occurs because there is a greater chance that the optimal phenotype is already occupied by a competitor in a more diverse community. More work is needed to examine how the distribution of different interaction types affects adaptation to environmental change in species-rich communities.

<!-- Conclusions -->
Our study gives insight to how local extinctions alter evolutionary constraints on remaining populations. In particular, it hints at a potential insidious effect of local extinctions that compromises the robustness of remaining populations to future environmental change. Our work also highlights some key challenges for predicting phenotypic evolution in the face of global change. First, the simplification of ecological communities may actually reduce the predictability of phenotypic evolution, because knowledge is required of both selective and genetic constraints, rather than potentially just genetic constraints in more complex systems. Second, many theoretical models of eco-evolutionary dynamics focus on phenotypic change in a single trait, yet our results highlight that the number of traits under selection may change with the community context. Importantly, we found that different species/guilds imposed different selection pressures. Knowing these hidden selection pressures is critical for prediction, because the trajectory of evolution will depend on the nature of change in the community context. We expect that a continued integration of adaptive landscapes and ecological networks will enhance our ability to predict the evolutionary consequences of changes in ecological communities.
